---
title: |
  \vspace*{\fill}
  \fontsize{36}{40}\selectfont \textbf {Predicting Median House Value in Boston}
author: |
  \fontsize{24}{28}\selectfont Seif H. Kungulio
date: |
  \fontsize{18}{22}\selectfont November 01, 2025
  \ \vspace*{\fill}
description: "A comprehensive analysis to predict median house values in Boston suburbs using various machine learning techniques."
categories: [Data Science, Machine Learning, R]
tags: [Regression, Random Forest, Data Preparation, Model Evaluation]
output: 
  pdf_document:
    latex_engine: xelatex
    toc: false
    highlight: espresso
    include:
      in_header: resources/header.tex
header-includes:
  # Font and color setup
  - \usepackage{xcolor}
  - \definecolor{brand}{HTML}{000000}
  - \definecolor{subbrand}{HTML}{212121}
  - \usepackage{titlesec}
  - \titleformat{\section}{\Large\bfseries\color{brand}}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries\color{subbrand}}{\thesubsection}{0.75em}{}
  - \usepackage[most]{tcolorbox}
  - \tcbset{colback=white, colframe=brand, coltitle=white, fonttitle=\bfseries}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{\textcolor{brand}{Predicting Median House Value in Boston}}
  - \fancyhead[R]{\textcolor{subbrand}{Seif H. Kungulio}}
  - \fancyfoot[C]{\thepage}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=6)

# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

#Install the following libraries if required
# List of required packages


required_pkgs <- c("MASS", "tidyverse","skimr", "caret","pROC","vip","xgboost",
                   "Matrix","pls","broom","janitor","car","yardstick","cowplot",
                   "ggrepel","patchwork", "e1071", "kableExtra", "glmnet", "randomForest", "ranger", "corrplot")

# Install and load each package
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

theme_set(theme_minimal(base_size = 12))
options(dplyr.summarise.inform = FALSE)
```

\pagenumbering{gobble}

\newpage
\thispagestyle{plain}
&nbsp;

\newpage
\pagenumbering{roman}
\tableofcontents

\newpage
\pagenumbering{arabic}


# **Business Understanding**
## Background & Problem Statement
Boston’s housing authority and urban developers face the challenge of accurately estimating property values across different neighborhoods. Property valuation depends on multiple interrelated factors—such as crime rate, accessibility to major roads, environmental quality, and local amenities—which can be difficult to model using traditional statistical techniques alone.

## Business Objectives
Build a predictive model to estimate median owner-occupied home value (medv) using neighborhood-level indicators. The goal is to support:

* neighborhood investment planning,
* redevelopment prioritization,
* and value forecasting under changing conditions.

## Success Criteria
* **Primary metric:** RMSE (lower is better)
* **Secondary metrics:** MAE and R²
* **Practical goal:** A model that generalizes well (low validation error) and offers interpretable drivers (feature importance / coefficients).

\newpage

# **Data Understanding**
## Data Collection
The dataset used in this analysis is the Boston Housing Dataset, which contains information about various attributes of houses in Boston suburbs. The dataset is publicly available and can be accessed through the `MASS` package in R.
```{r load-data}
data("Boston", package = "MASS")
boston.df <- Boston

#boston.df <- as_tibble(Boston) |>
#  clean_names()

glimpse(boston.df)
summary(boston.df$medv)
```

## Data Description
The dataset used in this project is the Boston Housing Dataset, originally published in the Harrison and Rubinfeld (1978) study and included in the MASS package in R. It contains 506 observations and 14 variables describing socioeconomic, demographic, environmental, and structural characteristics of housing in Boston suburbs. The dataset is commonly used for predictive modeling and regression analysis tasks, particularly for estimating housing values.

Each record represents a census tract in the Boston metropolitan area, with the target variable being the median value of owner-occupied homes (MEDV) expressed in $1000s.

The features cover a wide range of factors that influence property values — including crime rates, industrial activity, environmental quality, accessibility to major roads, and education levels. These variables provide a multi-dimensional view of the housing environment, making the dataset suitable for developing regression and machine learning models to predict housing prices.

## Data Dictionary

| Variable | Description                        | Data Type | Constraints/Rule |
|:---------|:-----------------------------------|:----------|:-----------------|
| **crim** | Per capita crime rate by town | Numeric | >= 0 |
| **zn**  | Proportion of residential land zoned for lots over 25,000 sq.ft. | Numeric | 0 <= **zn** <= 100 |
| **indus** | Proportion of non-retail business acres per town | Numeric | >= 0 |
| **chas** | Charles River dummy variable (1 if tract bounds river; 0 otherwise) | Integer | 0 or 1 |
| **nox** | Nitric oxides concentration (parts per 10 million) | Numeric | 0 < **nox** <= 1 |
| **rm** | Average number of rooms per dwelling | Numeric | > 0 |
| **age** | Proportion of owner-occupied units built prior to 1940 (%) | Numeric | 0 <= **age** <= 100 |
| **dis** | Weighted distances to five Boston employment centers | Numeric | > 0 |
| **rad** | Index of accessibility to radial highways | Integer | 1–24 |
| **tax** | Full-value property-tax rate per $10,000 | Numeric | > 0 |
| **ptratio** | Pupil-teacher ratio by town | Numeric | > 0 |
| **black** | (1000(Bk - 0.63)^2), where Bk is the proportion of Black residents by town | Numeric | >= 0 |
| **lstat** | Percentage of lower status of the population | Numeric | 0 <= **lstat** <= 100 |
| **medv** | Median value of owner-occupied homes in $1000s (Target variable) | Numeric | > 0 (typically capped at 50 in dataset) |

## Data Quality Assessment
```{r}
# Missing values
colSums(is.na(boston.df))

# Duplicates
sum(duplicated(boston.df))

# Quick profile
skim(boston.df)
```


## Exploratory Data Analysis (EDA)
### Target variable distribution
```{r}
ggplot(boston.df, aes(medv)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Median Home Value (medv)", x = "medv ($1000s)", y = "Count")
```

### Relationship with key predictors
```{r}
p1 <- ggplot(boston.df, aes(rm, medv)) + geom_point(alpha = 0.6) + geom_smooth(method="lm", se=FALSE) +
  labs(title="medv vs rm (rooms)")
p2 <- ggplot(boston.df, aes(lstat, medv)) + geom_point(alpha = 0.6) + geom_smooth(method="lm", se=FALSE) +
  labs(title="medv vs lstat")
p3 <- ggplot(boston.df, aes(crim, medv)) + geom_point(alpha = 0.6) + scale_x_log10() +
  geom_smooth(method="lm", se=FALSE) + labs(title="medv vs crim (log scale)")
p4 <- ggplot(boston.df, aes(nox, medv)) + geom_point(alpha = 0.6) + geom_smooth(method="lm", se=FALSE) +
  labs(title="medv vs nox")

(p1 | p2) / (p3 | p4)
```

### Correlation Overview

```{r}
num_df <- boston.df %>% select(where(is.numeric))
corr <- cor(num_df)

corrplot(corr, method = "color", type = "lower", tl.cex = 0.7)
```


\newpage

# **Data Preparation**
## Train/ Test Split
```{r}
set.seed(123)
idx <- createDataPartition(boston.df$medv, p = 0.8, list = FALSE)
train <- boston.df[idx, ]
test  <- boston.df[-idx, ]

dim(train); dim(test)
```

## Preprocessing Recipe
```{r}
preproc <- preProcess(
  train %>% select(-medv),
  method = c("center", "scale")
)

x_train <- predict(preproc, train %>% select(-medv))
x_test  <- predict(preproc, test %>% select(-medv))

y_train <- train$medv
y_test  <- test$medv
```


\newpage

# **Modeling**
## Evaluation Setup (Cross-Validation)
```{r}
ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 3
)

rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae  <- function(actual, pred) mean(abs(actual - pred))
r2   <- function(actual, pred) cor(actual, pred)^2
```

## Baseline Model (Mean Predictor)
```{r}
baseline_pred <- rep(mean(y_train), length(y_test))

baseline_metrics <- tibble(
  Model = "Baseline (Train Mean)",
  RMSE = rmse(y_test, baseline_pred),
  MAE  = mae(y_test, baseline_pred),
  R2   = r2(y_test, baseline_pred)
)

baseline_metrics
```

## Multiple Linear Regression
```{r}
lm_fit <- train(
  medv ~ ., data = train,
  method = "lm",
  trControl = ctrl
)

lm_pred <- predict(lm_fit, newdata = test)

lm_metrics <- tibble(
  Model = "Linear Regression",
  RMSE = rmse(y_test, lm_pred),
  MAE  = mae(y_test, lm_pred),
  R2   = r2(y_test, lm_pred)
)

lm_fit
lm_metrics
```

### Diagnostics (Residuals)
```{r}
res <- y_test - lm_pred

p_res1 <- ggplot(tibble(pred=lm_pred, res=res), aes(pred, res)) +
  geom_point(alpha=0.6) + geom_hline(yintercept = 0) +
  labs(title="Residuals vs Predictions", x="Predicted medv", y="Residual")

p_res2 <- ggplot(tibble(res=res), aes(sample=res)) +
  stat_qq() + stat_qq_line() +
  labs(title="Residual QQ Plot")

p_res1 | p_res2
```

## Regularized Regression (Elastic Net via glmnet)
```{r}
grid_glmnet <- expand.grid(
  alpha  = seq(0, 1, by = 0.25),
  lambda = 10^seq(-3, 1, length.out = 40)
)

glmnet_fit <- train(
  x = as.matrix(x_train),
  y = y_train,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = grid_glmnet
)

glmnet_pred <- predict(glmnet_fit, newdata = as.matrix(x_test))

glmnet_metrics <- tibble(
  Model = "Elastic Net (glmnet)",
  RMSE = rmse(y_test, glmnet_pred),
  MAE  = mae(y_test, glmnet_pred),
  R2   = r2(y_test, glmnet_pred)
)

glmnet_fit
glmnet_metrics
```

### Coefficient (Interpretable Drivers)
```{r}
best <- glmnet_fit$bestTune
best

final_glmnet <- glmnet_fit$finalModel
coef_mat <- coef(final_glmnet, s = best$lambda)

coef_df <- tibble(
  term = rownames(as.matrix(coef_mat)),
  estimate = as.numeric(as.matrix(coef_mat))
) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate)))

head(coef_df, 12)
```

## Random Forest (Non-linear + Interactions)
```{r}
set.seed(42)
rf_fit <- train(
  medv ~ ., data = train,
  method = "rf",
  trControl = ctrl,
  tuneLength = 8
)

rf_pred <- predict(rf_fit, newdata = test)

rf_metrics <- tibble(
  Model = "Random Forest",
  RMSE = rmse(y_test, rf_pred),
  MAE  = mae(y_test, rf_pred),
  R2   = r2(y_test, rf_pred)
)

rf_fit
rf_metrics
```

### Feature Importane (Random Forest)
```{r}
rf_final <- rf_fit$finalModel
imp <- randomForest::importance(rf_final)
imp_df <- tibble(
  Feature = rownames(imp),
  Importance = imp[,1]
) %>% arrange(desc(Importance))

head(imp_df, 12)

ggplot(imp_df %>% slice_max(Importance, n=12),
       aes(reorder(Feature, Importance), Importance)) +
  geom_col() +
  coord_flip() +
  labs(title="Top Feature Importances (Random Forest)", 
       x=NULL, 
       y="Importance")
```

\newpage

# **Evaluation**
## Compare Models on Test Set
```{r}
results <- bind_rows(baseline_metrics, 
                     lm_metrics, 
                     glmnet_metrics, 
                     rf_metrics) %>%
  arrange(RMSE)

results
```


## Prediction Quality Plot (Best Model)
```{r}
best_model <- results$Model[1]
best_model

best_pred <- switch(
  best_model,
  "Baseline (Train Mean)" = baseline_pred,
  "Linear Regression" = lm_pred,
  "Elastic Net (glmnet)" = glmnet_pred,
  "Random Forest" = rf_pred
)

plot_df <- tibble(
  actual = y_test,
  predicted = best_pred
)

ggplot(plot_df, aes(actual, predicted)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = paste("Actual vs Predicted (Best Model:", best_model, ")"),
    x = "Actual medv",
    y = "Predicted medv"
  )
```

## Error Analysis (Where the model misses)
```{r}
plot_df <- plot_df %>%
  mutate(error = actual - predicted,
         abs_error = abs(error)) %>%
  arrange(desc(abs_error))

head(plot_df, 10)

ggplot(plot_df, aes(error)) +
  geom_histogram(bins = 30) +
  labs(title="Prediction Error Distribution", 
       x="Actual - Predicted", 
       y="Count")
```


\newpage

# **Deployment**
```{r}
score_medv <- function(newdata, best_model_name = best_model) {
  # Ensure columns match training structure
  if (!all(names(train %>% select(-medv)) %in% names(newdata))) {
    stop("newdata must include all predictor columns used in training.")
  }
  newdata <- newdata %>% select(names(train %>% select(-medv)))

  pred <- switch(
    best_model_name,
    "Linear Regression" = predict(lm_fit, newdata = newdata),
    "Elastic Net (glmnet)" = {
      x_new <- predict(preproc, newdata)
      predict(glmnet_fit, newdata = as.matrix(x_new))
    },
    "Random Forest" = predict(rf_fit, newdata = newdata),
    rep(mean(y_train), nrow(newdata))
  )
  return(as.numeric(pred))
}

# Example: score the first 3 rows of the test set predictors
example_new <- test %>% select(-medv) %>% slice(1:3)
score_medv(example_new)
```


\newpage

# Conclusion


\newpage

# References



