---
title: "Median House Value in Boston"
output:
  flexdashboard::flex_dashboard:
    navbar:
    - href: "https://www.linkedin.com/in/seif-kungulio/"
      target: "_blank"
      icon: "fa-linkedin"
    - href: https://github.com/shkungulio
      target: "_blank"
      icon: "fa-github"
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

#Install the following libraries if required
library(shiny)
library(flexdashboard)
library(MASS)
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(scales)
library(tidyr)
library(tibble)

set.seed(42)

# Data
data("Boston", package = "MASS")
boston.df <- Boston

# Train the best model from the report: Random Forest
ctrl <- trainControl(method = "cv", number = 10)
rf_grid <- expand.grid(mtry = c(2, 4, 6, 8))

rf_fit <- train(
  medv ~ ., data = boston.df,
  method = "rf",
  trControl = ctrl,
  tuneGrid = rf_grid,
  importance = TRUE,
  ntree = 500
)

# Pre-compute variable importance for display
vi <- varImp(rf_fit, scale = TRUE)$importance |>
  tibble::rownames_to_column("feature") |>
  dplyr::arrange(dplyr::desc(Overall))

# Ranges for UI controls
rng <- boston.df |>
  dplyr::select(-medv) |>
  dplyr::summarise(dplyr::across(dplyr::everything(), list(min = min, max = max)))

get_min <- function(var) rng[[paste0(var, "_min")]]
get_max <- function(var) rng[[paste0(var, "_max")]]

fmt_money <- function(x) dollar(x * 1000)

```


Inputs {.sidebar}
-----------------------------------------------------------------------

Use the controls below to describe a Boston-area tract. Click **Predict** to estimate **MEDV** (median home value).

```{r}
hr()
sliderInput("rm", "Avg rooms per dwelling (rm)",
            min = floor(get_min("rm")), max = floor(get_max("rm")),
            value = median(boston.df$rm), step = 1)
hr()

sliderInput("lstat", "% lower status population (lstat)",
            min = get_min("lstat"), max = get_max("lstat"),
            value = median(boston.df$lstat), step = 0.1)
hr()

sliderInput("dis", "Distance to employment centers (dis)",
            min = get_min("dis"), max = get_max("dis"),
            value = median(boston.df$dis), step = 0.01)
hr()

sliderInput("nox", "Nitric oxides (nox)",
            min = get_min("nox"), max = get_max("nox"),
            value = median(boston.df$nox), step = 0.01)
hr()

actionButton("go", "Predict", class = "btn-primary")
hr()
```


Row
-----------------------------------------------------------------------

### Predicted MEDV

```{r}
newdata_reactive <- eventReactive(input$go, {

  # Start with median row
  base_row <- as.data.frame(
    lapply(boston.df[, -which(names(boston.df) == "medv")], median)
  )

  # Top 4 inputs from variable importance
  base_row$rm     <- input$rm
  base_row$lstat  <- input$lstat
  base_row$dis    <- input$dis
  base_row$nox    <- input$nox

  base_row
})

pred_medv <- eventReactive(input$go, {
  predict(rf_fit, newdata = newdata_reactive())
})

renderValueBox({
  req(pred_medv())
  valueBox(
    value = fmt_money(as.numeric(pred_medv())),
    caption = "Estimated median home value",
    icon = "fa-home",
    color = "primary"
  )
})
```
> The prediction demonstrates how multidimensional neighborhood factors translate into actionable price forecasts.

### **Quick Context**

```{r}
renderUI({
  req(pred_medv())

  p <- as.numeric(pred_medv())
  q <- quantile(boston.df$medv, probs = c(0.1, 0.5, 0.9))

  label <- dplyr::case_when(
    p < q[[1]] ~ "Lower-valued tract relative to the dataset",
    p < q[[3]] ~ "Mid-range tract relative to the dataset",
    TRUE ~ "Higher-valued tract relative to the dataset"
  )

  HTML(paste0(
    "<b>Model:</b> Random Forest (caret) — tuned via 10-fold CV<br/>",
    "<b>Best mtry:</b> ", rf_fit$bestTune$mtry, "<br/>",
    "<b>Interpretation:</b> ", label
  ))
})
```

Row
-----------------------------------------------------------------------

### **Average Room per Dwelling**

```{r}
renderPlot({
  req(newdata_reactive())

  ggplot(boston.df, aes(x = rm)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    geom_vline(xintercept = input$rm,
               color = "red",
               linewidth = 1.2) +
    labs(
      title = "Distribution of Average Rooms (rm)",
      subtitle = "Red line = Selected Input Value",
      x = "Average Rooms per Dwelling",
      y = "Count"
    ) +
    theme_test()
})
```
> Room count is the strongest valuation driver, with higher average rooms significantly increasing predicted home value.

### **% Lower Status Population**
```{r}
renderPlot({
  req(newdata_reactive())

  ggplot(boston.df, aes(x = lstat)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    geom_vline(xintercept = input$lstat,
               color = "red",
               linewidth = 1.2) +
    labs(
      title = "Distribution of % Lower Status Population (lstat)",
      subtitle = "Red line = Selected Input Value",
      x = "% Lower Status Population",
      y = "Count"
    ) +
    theme_test()
})
```
> Socioeconomic composition is the second most influential factor, with higher lstat values strongly reducing home prices.

Row
---

### **Distance to Employment Centers**
```{r}
renderPlot({
  req(newdata_reactive())

  ggplot(boston.df, aes(x = dis)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    geom_vline(xintercept = input$dis,
               color = "red",
               linewidth = 1.2) +
    labs(
      title = "Distribution of Distance to Employment Centers (dis)",
      subtitle = "Red line = Selected Input Value",
      x = "Weighted Distance",
      y = "Count"
    ) +
    theme_test()
})
```
> Accessibility moderately impacts pricing, with balanced proximity supporting stable valuations. The model captures the nonlinear relationship between commuting distance and housing demand.

### **Nitric Oxides Concentration**
```{r}
renderPlot({
  req(newdata_reactive())

  ggplot(boston.df, aes(x = nox)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    geom_vline(xintercept = input$nox,
               color = "red",
               linewidth = 1.2) +
    labs(
      title = "Distribution of Nitric Oxides Concentration (nox)",
      subtitle = "Red line = Selected Input Value",
      x = "Nitric Oxides Concentration",
      y = "Count"
    ) +
    theme_test()
})
```
> Environmental quality ranks among the top valuation drivers, with higher pollution levels negatively affecting home prices.


Model Insights {.tabset .tabset-fade}
-----------------------------------------------------------------------

### **Cross-validated Performance (Tuning)**

```{r}
renderTable({
  rf_fit$results |>
    dplyr::arrange(RMSE) |>
    dplyr::mutate(across(where(is.numeric), ~round(.x, 4)))
}, striped = TRUE, bordered = TRUE, hover = TRUE, spacing = "s")
```
> The tuned Random Forest achieves R² ≈ 0.89 and RMSE ≈ 3.10, explaining nearly 90% of price variance. Performance stability across folds confirms strong generalization and low overfitting risk.

### **Top Feature Importance**

```{r}
renderPlot({
  topn <- 10
  plot_df <- vi |>
    dplyr::slice_head(n = topn) |>
    dplyr::mutate(feature = reorder(feature, Overall))

  ggplot(plot_df, aes(x = Overall, y = feature)) +
    geom_col() +
    labs(
      title = paste0("Top ", topn, " drivers of MEDV (Random Forest importance)"),
      x = "Importance (scaled)",
      y = NULL
    )
})
```
> Structural capacity, socioeconomic composition, and environmental exposure dominate valuation impact. The importance ranking enhances interpretability while aligning with urban economic theory.

Row
-----------------------------------------------------------------------
The Random Forest model accurately predicts Boston median home values with ~89% variance explained. Structural housing characteristics, particularly average room count, are the strongest value drivers, followed by socioeconomic status and environmental quality indicators. The selected tract represents a mid-tier neighborhood with balanced accessibility and moderate environmental exposure. Overall, property valuation in Boston reflects a multidimensional interplay of structural, demographic, and environmental factors.
